$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-03 paper/paper.tex
Be careful not to confuse the reader with dots in the syntax

      \datatype
      {n : \N}
      {\Process_n : \Set}
      \; \textsc{Process}
    \end{mathpar}
  \end{subfigure}
  \begin{subfigure}[t]{.2\textwidth}
    \begin{equation*}
      \begin{split}
        P,Q ::=      & \; \PO                \\ 
                    |& \; (\new{}x)  P       \\ 
                    |& \; \comp{P}{Q}        \\ 
                    |& \; \recvp{x}{y}. \; P \\ 
                    |& \; \sendp{x}{y}. \; P 
      \end{split}
    \end{equation*}
    \caption{Ill-scoped grammar with names}
    \label{fig:syntax-names}
  \end{subfigure}
  \hspace{.1\textwidth}
  \begin{subfigure}[t]{.7\textwidth}
    \begin{equation*}
      \begin{split}
        \Process_n ::=& \; \PO_n               &&\text{(process termination)}   \\ 
        |& \; \new{} \Process_{\suc n}         &&\text{(restriction)}          \\ 
        |& \; \comp{\Process_n}{\Process_n}    &&\text{(parallel composition)} \\ 
        |& \; \recv{\Var_n}{}\Process_{\suc n} &&\text{(input)}                \\ 
        |& \; \send{\Var_n}{\Var_n}\Process_n  &&\text{(output)}                  
      \end{split}
    \end{equation*}
    \caption{Well-scoped grammar with type-level de Bruijn indices}
    \label{fig:syntax}
  \end{subfigure}
  \caption{}
\end{figure}

Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y}. \; P$ and $\sendp{x}{y}. \; P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x) \; P$ and input $\recvp{x}{y}. \; P$ both introduce bound names in $P$.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: inserting a new variable into a context means proving that its name differs from all other variable names in context.
Furthermore, names do not guarantee that the syntax well-scoped.
Instead, we use de Bruijn indices \cite{deBruijn1972}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
  \begin{equation*}
    \begin{split}
      & (\new{}x ) && (\comp {\recvp{x}{y} && . \; \sendp{y}{\texttt{"Hello world!"}} . \; \PO} {(\new{} y) && (\sendp{x}{y} . \; \recvp{y}{z} && . \; \PO)})
      & \text{names}
      \\
      & \new{} && (\comp {\recv{0} && . \; \send{0}{\texttt{"Hello world!"}} . \; \PO} {\new{} && (\send{1}{0} . \; \recv{0} && . \; \PO)})
      & \text{de Bruijn}
      \\
    \end{split}
  \end{equation*}
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.

>>>>

      \datatype
      {n : \N}
      {\Process_n : \Set}
      \; \textsc{Process}
    \end{mathpar}
  \end{subfigure}
  \begin{subfigure}[t]{.2\textwidth}
    \begin{equation*}
      \begin{split}
        P,Q ::=      & \; \PO                \\ 
                    |& \; (\new{}x) \; P       \\ 
                    |& \; \comp{P}{Q}        \\ 
                    |& \; \recvp{x}{y} P \\ 
                    |& \; \sendp{x}{y} P 
      \end{split}
    \end{equation*}
    \caption{Ill-scoped grammar with names}
    \label{fig:syntax-names}
  \end{subfigure}
  \hspace{.1\textwidth}
  \begin{subfigure}[t]{.7\textwidth}
    \begin{equation*}
      \begin{split}
        \Process_n ::=& \; \PO_n               &&\text{(process termination)}   \\ 
        |& \; \new{} \Process_{\suc n}         &&\text{(restriction)}          \\ 
        |& \; \comp{\Process_n}{\Process_n}    &&\text{(parallel composition)} \\ 
        |& \; \recv{\Var_n}{}\Process_{\suc n} &&\text{(input)}                \\ 
        |& \; \send{\Var_n}{\Var_n}\Process_n  &&\text{(output)}                  
      \end{split}
    \end{equation*}
    \caption{Well-scoped grammar with type-level de Bruijn indices}
    \label{fig:syntax}
  \end{subfigure}
  \caption{}
\end{figure}

Process $\PO$ denotes the terminated process, where no operations can further occur.
Process $(\new{}x)\; P$ creates a new channel $x$ bound with scope $P$.
Process $\comp{P}{Q}$ is the parallel composition of processes $P$ and $Q$.
Processes $\recvp{x}{y} P$ and $\sendp{x}{y} P$ denote respectively the input and output processes of a variable $y$ over a channel $x$, with continuation $P$.

Scope restriction $(\new{}x) \; P$ and input $\recvp{x}{y} \; P$ both introduce bound names in $P$.
In order to mechanise the \picalc{} syntax in Agda, we need to deal with bound names in continuation processes.
Names are cumbersome to mechanise: inserting a new variable into a context means proving that its name differs from all other variable names in context.
Furthermore, names do not guarantee that the syntax well-scoped.
Instead, we use de Bruijn indices \cite{deBruijn1972}, where a natural number $n$ (aka \emph{index}) is used to refer to the variable introduced $n$ binders ago --- hence binders no longer introduce names.
\begin{example}
  \begin{equation*}
    \begin{split}
      & (\new{}x ) && (\comp {\recvp{x}{y} && \; \sendp{y}{a} \; \PO} {(\new{} y) && (\sendp{x}{y} \; \recvp{y}{z} && \; \PO)})
      && \text{names}
      \\
      & \new{} && (\comp {\recv{0} && \; \send{0}{a} \; \PO} {\new{} && (\send{1}{0} \; \recv{0} && \; \PO)})
      && \text{de Bruijn}
      \\
    \end{split}
  \end{equation*}
\end{example}
That is, terms at different \emph{depths} use different indices to refer to the same binding.
